I feel obliged to say special words of thank to Yannick Forster, who advised this thesis and the development of the framework presented in this
thesis.  In many weekly meetings, over the course of one year, we discussed how the framework could look like.  Over the course of the development, he
also helped me to organise myself.  I am thankful that he provided a {\LaTeX} template for this thesis and that he proof read my texts many times.

I want to thank my supervisor Prof.~Smolka, who introduced me to the field of functional programming and computational logic.  I always appreciated
his passion in his lectures.  Without him, I would not have found the field of computational logic and formal verification.  I also recognise the help
of Kathrin Stark, who helped me getting into this topic, in particular because I missed the first three weeks of \textit{Introduction to Computational
  Logic}.  I want to thank Prof.~Smolka for giving me the opportunity to write my thesis at his chair.  I really appreciate the atmosphere together
with motivated PhD students, especially with Kathrin Stark, Fabian Kunze, and Yannick Forster.

I express gratitude to Prof.~Hermanns, who mentored me in the computer science undergraduate Honours Program of Saarland University.  I am glad that
he is the second reviewer of this thesis.

Lastly, I want to thank my parents for supporting me in every imaginable way.  Without my parents, this thesis would not exists, me neither.


%%% Local Variables:
%%% TeX-master: "thesis"
%%% mode: LaTeX
%%% End:
