\chapter{Simple Machines}
\label{chap:compound}

In this chapter, we build and verify simple machines using the primitive machines from Chapter~\ref{chap:basic}, the operators developed in
Chapter~\ref{chap:combining}, and the tapes-lift from Chapter~\ref{chap:lifting}.  The machines in this chapter will be useful in the next chapter.
We do not use the alphabet-lift yet.  First we show how we prove correctness and termination of machines from now on.

When we prove $M \Realise R$ for a machine $M$ and its correctness relation $R$, we first find a relation $R'$ that $M$ realises.  We derive this
relation by applying the correctness lemmas of the control-flow operators, lifts, and concrete machines.  This process is mechanical and does in
general not depend on arbitrary choices (with a few exceptions).  The derived relation respects the ``structure'' of the machine.  For example, the
relation of a sequential composition is the relational composition of two relations.  Using the monotonicity Lemma~\ref{lem:Realise_monotone}, it
remains to show $R' \subseteq R$.  Because the structure of $R'$ respects the structure of $M$, the proof of the inclusion also follows its structure.
For example, the relation for a conditional is $(R_1 \at \true \circ R_2) \cup (R_1 \at \false \circ R_3)$.  From that it follows that we do a
case-distinction for both branches in the proof.  Note that we do not have to reason about machine states at all, because the correctness relations
are only relations between tapes and labels.

% So the main part of the proof is showing the inclusion.  The ``canonical relation'' of Section~\ref{sec:canonical} is for that reason not suited for
% the choice of $R'$.

When $M$ always terminates in constant time $k$, we show $M \RealiseIn{k} R$ instead.  Using the monotonicity Lemma~\ref{lem:RealiseIn_monotone}, we
can prove correctness and constant time at once.  For non-constant running time, we show $M \TerminatesIn T$ for a running time relation~$T$.  For
that, we use the dual approach and apply the anti-monotonicity Lemma~\ref{lem:TerminatesIn_monotone}.


\section{$\Nop$}
\label{sec:Nop}
\setCoqFilename{ProgrammingTuringMachines.TM.Compound.Multi}%


Using the tapes-lift (Definition~\ref{def:LiftTapes}) and $\MS{Null}$ (Definition~\ref{def:Null}), it is easy to define an $n$-tape machine
$\Nop : \TM_\Sigma^n$ that does nothing.  Asperti and Ricciotti~\cite{asperti2015} define this machine directly:
\begin{definition}[$\Nop$][Nop]
  $\Nop \defop \LiftTapes{\MS{Null}}{\nil}$.
\end{definition}
Note that because $\MS{Null}$ is a 0-tape machine, and $\Nop$ is supposed to be an $n$-tape machine, the index-vector must be the vector
$\nil : \Fin_n^0$.

\begin{lemma}[Correctness of $\Nop$][Nop_Sem]
  \label{lem:Nop_Sem}
  $\Nop \RealiseIn0 NopRel$ with $NopRel := \lambda t~t'.~t'=t$.
\end{lemma}
\begin{proof}
  We apply the monotonicity Lemma~\ref{lem:RealiseIn_monotone} of $\RealiseIn\cdot$, the correctness Lemma~\ref{lem:Null_Sem} of $\MS{Null}$, and the
  correctness Lemma~\ref{lem:LiftTapes_Realise} of the tapes-lift.  It remains to show:
  \[
    \left( \lambda t~t'.~ NullRel~t~t' \land \left(\forall i:\Fin_n.~i \notin \nil \rightarrow t'[i] = t[i] \right)\right)
    \subseteq NopRel
  \]
  Let $t,t' : \Tape_\Sigma^n$.  To show the equality $t'=t$ we show $t'[i]=t[i]$ for all $i:\Fin_n$.  This follows with the equality part of the
  relation $\LiftTapes{NullRel}{\nil}$, since $i \notin \nil$.
\end{proof}

Note that the correctness relation of $\Nop$ can also be expressed using the identity relation $Id$:
\[
  NopRel \equiv Id.
\]
We have the convention to define relations of concrete machines in $\lambda$-notation, i.e.\ not using relational operators.  Also note that the tape
$t'$ is, per convention, always on the left side of the equality.  These conventions make rewriting of tapes uniform; therefore, rewriting of tapes
can be automated in Coq.

\section{$\MS{WriteString}$}
\label{sec:WriteString}
\setCoqFilename{ProgrammingTuringMachines.TM.Compound.WriteString}%

The machine $\MS{WriteString}~d~str$ writes a fixed string $str:\List(\Sigma)$ in the direction $d$.  It is defined by recursion over the string:
\begin{definition}[$\MS{WriteString}$][WriteString]
  \begin{alignat*}{3}
    &\MS{WriteString}~d~&&(\nil)         &~:=~& \Nop \\
    &\MS{WriteString}~d~&&(s \cons \nil) &~:=~& \MS{Write}~s \\
    &\MS{WriteString}~d~&&(s \cons str') &~:=~& \MS{WriteMove}~s~d \Seq \MS{WriteString}~d~str'.
  \end{alignat*}
\end{definition}

Note that this is our only machine we define per recursion.  The way we prove correctness in constant time (depending on the length of $str$) is still
the same.

The machine writes all symbols of the string $str$ to the tape and moves in the tape in direction $d$ after each (but the last) symbol.  When it
terminates, the head of the tape is under the last written symbol, which is the last symbol of $str$.  It terminates in constant time, after
$2\cdot\length{str}-1$ steps.

The derived relation for $\MS{WriteString}$ is also defined per recursion over the string:
\begin{lemma}[][WriteString_fix_Sem]
  $\MS{WriteString}~d~str \RealiseIn{2 \cdot \length{str} - 1} R'~d~str$ with
  \begin{alignat*}{3}
    &\MS{R'}~d~&&(\nil        ) &~:=~& NopRel \\
    &\MS{R'}~d~&&(s \cons \nil) &~:=~& DoActRel(\Some{s}, \MS{N}) \\
    &\MS{R'}~d~&&(s \cons str') &~:=~& DoActRel(\Some{s}, d) \circ \MS{R'}~d~str'.
  \end{alignat*}
\end{lemma}
\begin{proof}
  By induction on $str:\List(\Sigma)$, using the monotonicity Lemma~\ref{lem:RealiseIn_monotone}, the correctness of $\Nop$ (Lemma~\ref{lem:Nop_Sem}),
  the correctness of $\MS{Write}$ and $\MS{WriteMove}$ (which are defined using $\MS{DoAct}$; Lemma~\ref{lem:DoAct_Sem}), and correctness of
  sequential composition for constant time (Lemma~\ref{lem:Seq_RealiseIn}).
\end{proof}

We define the actual relation of $\MS{WriteString}$ in terms of a function on tapes that is also defined by recursion over $str$:
\begin{lemma}[Correctness of $\MS{WriteString}$][WriteString_Sem]
  \label{lem:WriteString_Sem}
  Let $d:\Move$ and $str:\List(\Sigma)$.
  \[ \MS{WriteString}~d~str \RealiseIn{2\cdot\length{str}-1} WriteStringRel~d~str \]
  with
  $WriteStringRel~d~str := \lambda t~t'.~t' = writeStringFun~d~t~str$ and
  \begin{alignat*}{3}
    &writeStringFun~d~t~&&(\nil        ) &~:=~& t \\
    &writeStringFun~d~t~&&(s \cons \nil) &~:=~& \MS{wr}~t~(\Some{s}) \\
    &writeStringFun~d~t~&&(s \cons str') &~:=~& writeStringFun~d~\bigl(\MS{doAct}~t~(\Some{s}, d)\bigr)~str'
  \end{alignat*}
\end{lemma}
\begin{proof}
  We apply the monotonicity Lemma~\ref{lem:RealiseIn_monotone} and have to show:
  \[
    R'~d~str \subseteq WriteStringRel~d~str
  \]
  This can be shown by induction on $str$.
\end{proof}

Note that we could as well use the $\MS{Mirror}$ operator instead of parametrising the machine $\MS{WritingString}$ over the direction.  In this
particular example the parametrising approach seems to be easier.

\section{$\MS{MovePar}$}
\label{sec:MovePar}
\setCoqFilename{ProgrammingTuringMachines.TM.Compound.Multi}%

The two-tape machine $\MS{MovePar}~d_0~d_1$ combines two $\MS{Move}$ machines.  It first moves the $0$th tape in direction $d_0$ and after that the
$1$st tape in direction $d_1$.\footnote{To avoid confusion with zero-based indices used throughout this thesis, we write ``the $0$th or $1$st tape'',
  instead of ``the first or second tape.''}
\begin{definition}[$\MS{MovePar}$][MovePar]
  \label{def:MovePar}
  $\MS{MovePar}~d_0~d_1 \defop \LiftTapes{(\MS{Move}~d_0)}{\Vector{0}} \Seq \LiftTapes{(\MS{Move}~d_1)}{\Vector{1}}$.
\end{definition}
\begin{lemma}[Correctness of $\MS{MovePar}$][MovePar_Sem]
  \label{lem:MovePar_Sem}
  $\MS{MovePar}~d_0~d_1 \RealiseIn3 MoveParRel~d_0~d_1$ with
  \[
    MoveParRel~d_0~d_1 := \lambda t~t'.~t'[0]=\MS{mv}~d_0~t[0] ~\land~ t'[1]=\MS{mv}~d_1~t[1]
  \]
\end{lemma}
\begin{proof}
  We have to show:
  \[
    \LiftTapes{(DoActRel(\None,d_0))}{\Vector{0}} \circ
    \LiftTapes{(DoActRel(\None,d_1))}{\Vector{1}} \subseteq
    MovePairRel~d_0~d_1
  \]
  We assume tape vectors $t, t', t'' : \Tape_\Sigma^2$, such that $(\LiftTapes{(DoActRel(\None,d_0))}{\Vector{0}})~t~t'$ and \\
  $(\LiftTapes{(DoActRel(\None,d_1))}{\Vector{1}})~t'~t''$.  We have to show $t''[0]=\MS{mv}~d_0~t[0]$ and $t''[1]=\MS{mv}~d_1~t[1]$.  By definition,
  we know $t'[0]=\MS{mv}~d_0~t[0]$ and $t'[1]=t[1]$ (because $1 \notin \Vector{0}$).  We also know $t''[1]=\MS{mv}~d_1~t'[1]$ and $t''[0]=t'[0]$
  (because $0 \notin \Vector{1}$).  The goal follows trivially.
\end{proof}
Note that this kind of proof is very mechanical: We only need to unfold the relations and rewrite tapes.  Indeed, these steps are automated in Coq.
Thus, we also do not present more proofs of this kind on paper.

\section{$\MS{CopySymbols}$}
\label{sec:CopySymbols}
\setCoqFilename{ProgrammingTuringMachines.TM.Compound.CopySymbols}%

The machine $\MS{CopySymbols}~h : \TM_\Sigma^2$, where $h:\Sigma\to\Bool$, is a compound machine involving a $\While$-loop.  It reads a symbol on tape
$0$, writes it to tape $1$, and moves both tapes to right, until the read symbol satisfies $h$.  If there was no current symbol on tape $0$, it also
terminates.

We first define the machine for the step.  Since we want to apply the $\While$ operator on the step machine, it must be labelled over
$\Option(\Unit)$.  $\Some\unit$ means to break out of the loop and $\None$ means to repeat the loop.
\begin{definition}[$\MS{CopySymbolsStep}$][CopySymbols_Step]
  \label{CopySymbols_Step}
\begin{lstlisting}[style=semicoqstyle]
$\MS{CopySymbolsStep}~h :=$
  Switch$(\LiftTapes{\MS{Read}}{\Vector{0}})$
       $(\lambda (s:\Option(\Sigma)).$
          $\MS{match}~s$
          [$\Some{x}$=>
            $\MS{if}~h(x)$
            $\pthen$ $\Return{\bigl(\LiftTapes{(\MS{Write}~x)}{\Vector{1}}\bigr)}{\Some\unit}$ 
            $\pelse$ $\Return{\bigl(\LiftTapes{(\MS{Write}~x)}{\Vector{1}} \Seq \MS{MovePar}~\MS{R}~\MS{R}\bigr)}{\None}$ 
          |$\None$ => $\Return{\Nop}{\Some\unit}$ 
          ])
\end{lstlisting}
\end{definition}

Note that ``$\MS{match}~[~\cdots~]$'' denotes pattern matching our type theory.

\begin{lemma}[Correctness of $\MS{CopySymbolsStep}$][CopySymbols_Step_Sem]
  \label{lem:CopySymbols_Step_Sem}
  ~
  \[
    \MS{CopySymbolsStep} \RealiseIn{7} CopySymbolsStepRel
  \]
  with
  \small
  \begin{align*}
    &CopySymbolsStepRel := \lambda t~(l, t').~\\
    &\quad\begin{cases}
      t'[0] = t[0]           \land t'[1]=\MS{wr}~t[1]~\Some{x}              \land l=\Some\unit & \MS{current}~t[0]=\Some{x} \land       h(x) \\
      t'[0] = \MS{mv}~\MS{R}~t[0] \land t'[1]=\MS{doAct}~t[1]~(\Some{x}, \MS{R}) \land l=\None & \MS{current}~t[0]=\Some{x} \land \lnot h(x) \\
      t' = t \land l=\Some\unit                                                                & \text{else}
    \end{cases}
  \end{align*}
\end{lemma}
\begin{proof}
  Mechanical, with case-analysis over $\MS{current}~t[0]$.
\end{proof}

We define $\MS{CopySymbol}$ by applying the $\While$ operator to $\MS{CopySymbolsStep}$:
\begin{definition}[$\MS{CopySymbols}$][CopySymbols]
  \label{def:CopySymbols}
  $\MS{CopySymbols}~h := \While(\MS{CopySymbolsStep}~h)$.
\end{definition}

The correctness of $\MS{CopySymbols}$ can be expressed using a recursive function on tapes:
\begin{lemma}[Correctness of $\MS{CopySymbols}$][CopySymbols_Realise]
  \label{lem:CopySymbols_Realise}
  $\MS{CopySymbols}~h \Realise CopySymbolsRel~h$
  with $CopySymbolsRel~h := \lambda t~t'.~t' = copySymbolsFun~h~t$ and
  {
    \small
    \begin{align*}
      &copySymbolsFun~h~t :=\\
      &\quad\begin{cases}
        \Vector{t[0];~ \MS{wr}~t[1]~\Some{x}}                                         & \MS{current}~t[0]=\Some{x} \land h(x) \\
        copySymbolsFun~h~\Vector{\MS{mv}~R~t[0];~ \MS{doAct}~t[1]~(\Some{x}, \MS{R})} & \MS{current}~t[0]=\Some{x} \land \lnot h(x) \\
        t                                                                             & \MS{current}~t[0]=\None
      \end{cases}
    \end{align*}
  }
  Note that the function $copySymbolsFun$ is not structural recursive.  It terminates because tapes have only finitely many symbols.
\end{lemma}
\begin{proof}
  To show: $WhileRel~CopySymbolsStepRel \subseteq CopySymbolsRel$.  By $\While$-induction (Lemma~\ref{lem:WhileInduction}).
\end{proof}

We observe that the running time of $\MS{CopySymbols}$ only depends on the $0$th tape.  Therefore, we define a function
$copySymbolsSteps : \Tape_\Sigma \to \Nat$ that overestimates the number of steps needed for the loop, depending on the $0$th tape.  Note that
$\While$ requires one additional step for each repeat of the loop.
\begin{lemma}[Running time of $\MS{CopySymbols}$][CopySymbols_Terminates]
  $\MS{CopySymbols} \TerminatesIn CopySymbolsT$ with \\
  $CopySymbolsT := \lambda t~k.~copySymbolsSteps(t) \leq k$ and
  \begin{align*}
    &copySymbolsSteps (t) := \\
    &\quad\begin{cases}
      8 + copySymbolsSteps(\MS{mv}~\MS{R}~t) & \MS{current}~t=\Some{x} \land \lnot h(x) \\
      8                                      & \text{otherwise}
    \end{cases}
  \end{align*}
\end{lemma}
\begin{proof}
  We have to show $CopySymbolsT \subseteq WhileT~CopySymbolsStepRel~(\lambda \_~k.~7 \leq k)$, using the co-induction
  Lemma~\ref{lem:WhileCoInduction}.  Let $copySymbolsSteps~t[0] \leq k$.  We choose $k_1 := 7$.  We have two cases.
  \begin{enumerate}
  \item We assume $CopySymbolsStepRel~t~(\Some\unit, t')$.  Therefore, we know that either $\MS{current}~t[0]=\Some{x}$ with $h(x)=\true$, or
    $\MS{current}~t[0]=\None$.  In both cases, we have $copySymbolsSteps~t[0] = 8$.  Thus, we have:
    $$k_1 \leq copySymbolsSteps~t[0] = 8 \leq k$$
  \item We assume $CopySymbolsStepRel~t~(\None, t')$.  Therefore, we have $\MS{current}~t[0]=\Some{x}$ with $h(x)=\false$, and
    $t'[0]=\MS{mv}~\MS{R}~t[0]$.  Then, we have:
    % \[ ... \] doesn't work here. $$ ... $$ doesn't place \qed at the right spot.
    \begin{equation*}
      1+k_1+copySymbolsSteps~t'[0] = copySymbolsSteps~t[0] \leq k
    \end{equation*}
  \end{enumerate}
\end{proof}
% Note that we see in the proof that we could replace the $8$s in the non-recursive parts of the running time functions with $7$s.

Using the $\MS{Mirror}$ operator, we can define a machine $\MS{CopySymbolsL}$ that copies and goes to the left instead.  We also have to ``mirror''
the correctness relations and their respective functions.  We do not repeat the definitions here.
\begin{definition}[$\MS{CopySymbolsL}$][CopySymbols_L]
  $\MS{CopySymbolsL}~h := \MS{Mirror}(\MS{CopySymbols}~h).$
\end{definition}

\section{$\MS{MoveToSymbol}$}
\label{sec:MoveToSymbol}
\setCoqFilename{ProgrammingTuringMachines.TM.Compound.MoveToSymbol}%

\enlargethispage{0.5cm}

We can define a machine $\MS{MoveToSymbol}~h~f : \TM_\Sigma^1$, where $h:\Sigma\to\Bool$ and $f:\Sigma\to\Sigma$.  This machine behaves similar as
$\MS{CopySymbols}~h$.  Instead of copying the symbols from one tape to another tape, it ``translates'' the symbols it reads, until it reads a symbol
that satisfies the boolean predicate $h$.  We leave out the correctness and running time statements, as they can be derived from the statements about
$\MS{CopySymbols}$ above.

\begin{definition}[$\MS{MoveToSymbol}$][MoveToSymbol]
  ~
\begin{lstlisting}[style=semicoqstyle]
$\MS{MoveToSymbolStep}~h~f :=$
  Switch$(\MS{Read})$
       $(\lambda (s:\Option(\Sigma)).~\MS{match}~s$
          [$\Some{x}$=> 
            if $h(x)$
            $\pthen$ $\Return{\bigl(\MS{Write}~(f~x)       \bigr)}{\Some\unit}$ 
            $\pelse$ $\Return{\bigl(\MS{WriteMove}~(f~x)~R \bigr)}{\None}$ 
          |$\None$ => $\Return{\Nop}{\Some\unit}$ 
          ])
\end{lstlisting}
  \begin{align*}
    \MS{MoveToSymbol }~h~f &:= \While(\MS{MoveToSymbolStep}~h~f) \\
    \MS{MoveToSymbolL}~h~f &:= \MS{Mirror}(\MS{MoveToSymbol}~h~f)
  \end{align*}
\end{definition}



%%% Local Variables:
%%% TeX-master: "thesis"
%%% End: